Generating Correlated Data for Omics Simulation

### 2025-06-17 manuscript/paper.qmd
Narrower margins for pdf
Finalizing revisions

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Department of Genetics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
  warning: false
format:
  docx:
    fig-dpi: 800
  html:
    toc: true
    toc-expand: true
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
bibliography: references.bib
---
## Abstract

Simulation of realistic omics data is a key input for benchmarking studies that help users obtain optimal computational pipelines.
Omics data involves large numbers of measured features on each samples and these measures are generally correlated with each other.
However, simulation too often ignores these correlations, perhaps due to computational and statistical hurdles of doing so.
To alleviate this, we describe three approaches for efficiently generating omics-scale data with correlated measures which mimic real datasets.
These approaches are all based on a Gaussian copula approach with a covariance matrix that decomposes into a diagonal part and a low-rank part.
We use these approaches to demonstrate the importance of including correlation in two benchmarking applications.
First, we show that variance of results from the popular DESeq2 method increases when dependence is included.
Second, we demonstrate that CYCLOPS, a method for inferring circadian time of collection from transcriptomics, improves in performance when given gene-gene dependencies in some circumstances.
We provide an R package, dependentsimr, that has efficient implementations of these methods and can generate dependent data with arbitrary marginal distributions, including discrete (binary, ordered categorical, Poisson, negative binomial), continuous (normal), or with an empirical distribution.

@@

Modern techniques, including high-throughput sequencing, produce more data than ever before.
To determine the optimal computational analysis methods for these data, benchmarks are often performed using simulated data.
This simulated data needs to closely match realistic data in order for benchmarking to meaningful.
An often neglected aspect of this is that measurements of different values are often correlated or dependent upon each other.
Two possible reasons for this neglect could be that there is a lack of guidelines on how to produce such data and also that methods to produce it are computationally expensive to run.
We describe here three related methods that are both conceptually relatively simple and also highly computationally efficient.
We demonstrated these on two applications which show how inclusion of these dependencies can affect the results of benchmarking.
Lastly, we provide a software package to act as reference implementations of these.

@@

Using a Gaussian copula [@Nelsen1998-nq] approach (also referred to as NORTA, for "normal to anything" [@norta]), the marginal (univariate) distributions can have realistic forms.
These solutions operate by taking a real, reference dataset and producing simulations that mimic it.
However, a major challenge when simulating datasets is choice of the copula, which is a mathematical object that encodes the dependence between two or more variables.
For Guassian copulas, this is a choice a covariance matrix.
However, the sample covariance matrix of a reference dataset is a poor choice and present three alternative choices.

There is a large body of literature on determining covariance matrices for various purposes.
Shrinkage methods [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] improve estimates and guarantee that the estimated matrix is well-conditioned (and hence invertible).
Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] produce a matrices where many of the pairwise correlations are zero.
Similarly, other methods produce a sparse estimate of the inverse of the covariance matrix [@sparse_inverse_covariance], which imparts pairwise independence of some variables conditional on all the others.

Copula approaches are flexible and have a long history [@Sklar] of use in fields such as economics [@MacKenzie_copula] 
More recently, copula approaches have been used for modelling or classification in multi-omics fields [@He_copula; @Ma_copula] and metagenomics [@Deek_microbiome].
Projects simulating omics with copulas include SPsimSeq [@Assefa2020] and scDesign2 [@Sun2021].
Some methods have employed the "vine copula" approach in the context of single-cell or spatial omics simulation [@fuetterer2019constructing; @scDesign3].
Vine copulas provide an alternative approach for choosing a copula which decomposes it into a set of 2-dimensional dependencies, however it comes at significant computational cost when done at the omics-scale.
SeqNet uses instead a network approach for gene dependence [@SeqNet].

Here, we describe three different strategies for determining covariance matrices, all of which enable efficient generation of simulated data due to following a specific form.
Specifically, these all decompose into a diagonal part and a low-rank part, which speeds up generation of random vectors.
First, the 'PCA' method uses principal component analysis (PCA) and picks the low-rank part such that the simulated data has the same variance in the top $k$ PCA components of the reference dataset.
Second, the 'spiked Wishart' method fits $k$ components such that simulations with the same number of samples as the reference dataset will have, on average, the same PCA component variances as the reference.
Unlike 'PCA', these variances are computed with respect to the simulated data's own PCA rather than using the PCA weights of the reference dataset.
Third, the 'corpcor' method builds off the popular R library `corpcor` [@SchaferStrimmer2005; @OpgenRheinStrimmer2007], which implements a James-Stein type shrinkage estimator for the covariance matrix as a linear interpolation of the sample covariance matrix and a diagonal matrix.
No method exactly replicates the reference dataset's properties, indicating room for future research, but all improve upon the common approach of assuming independence.
We implement this in an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors), and empirical (for ordinal data) marginal distributions.

@@

We start with a reference dataset $X$ given by a $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real dataset of interest.
Our goal is to be able to generate simulated datasets that look the same as this reference dataset, on average and when $n$ samples were simulated.
To do this, we need to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
We refer to these features as genes for simplicity.

Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
See @fig-intro a for a high-level overview.
This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
In omics, choosing the covariance matrix $\Sigma_{\text{sim}}$ to use during simulation presents a challenge since $n < p$.
The most obvious choice is to use the sample covariance matrix $\widehat \Sigma_{\text{ref}}$ of the reference dataset (possibly after some transformation of the data).
However, using $\widehat \Sigma_{\text{ref}}$ is a poor choice for $\Sigma_{\text{sim}}$.
Indeed, we consider a choice of $\Sigma_{\text{sim}}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{\text{sim}}$ have sample covaraince matrices $\widehat \Sigma_{\text{sim}}$ that are similar to $\widehat \Sigma_{\text{ref}}$.
Counter-intuitively, data simulated using $\Sigma_{\text{sim}} := \widehat \Sigma_{\text{ref}}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{\text{ref}}$, see @fig-intro b.

```{r}
#| label: fig-intro
#| fig-cap: "(a) Schematic of our Gaussian copula-based approach. The goal is to pick $\\Sigma_{\\text{sim}}$ so that $\\widehat \\Sigma_{\\text{sim}} \\approx \\widehat \\Sigma_{\\text{ref}}$ while using a special, highly efficient form. (b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{\\text{ref}}$ of reference data compared to $\\widehat \\Sigma_{\\text{sim}}$ from simulated data using $\\Sigma_{\\text{sim}} = \\widehat \\Sigma_{\\text{ref}}$. 100 genes and 12 samples were used for both reference and the 20 simulated datasets. Both reference and simulated data follow a multivariate normal distribution. Eigenvalues are numbered from largest to smallest. (c-d) Comparison of (c) run time and (d) memory usage for our methods (pca, wishart, coprcor), SPsimSeq, multivariate normal sampling (using the mvrnorm function from the MASS package), and vine copula (implemented by the rvinecopulib R package). 100 samples were generated with a varying number of genes. A mouse cortex dataset GSE151564 was randomly subsetted to have the required number of genes to be used as the reference data set. Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts."
#| fig-width: 8
#| fig-height: 6

@@

One reason this occurs is that $\widehat \Sigma_{\text{ref}}$ is low rank, specifically at most rank $n - 1$, which is much less than $p$ in typical omics studies.
This means that any data generated using $\Sigma_{\text{sim}} = \widehat \Sigma_{\text{ref}}$ will lie on an $n-1$-dimensional hyperplane.
Intuitively, this means that the entire probability distribution of potentially $p$ dimensional data has been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference dataset.
In contrast, we know that omics contains at least some level of random noise that is essentially independent from one gene to the next.
For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
We therefore need to "inject" some level of additional randomness that is independent between genes on top of the correlations captured by $\widehat \Sigma_{\text{ref}}$

We present three methods - to be called PCA, spiked Wishart, and corpcor - for making the choice of $\Sigma_{\text{sim}}$
All three take the form
$$ \Sigma_{\text{sim}} = D^2 + UW^2U^T $$
where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{\text{ref}}$.
An additional advantage of this special form is that data may be generated very efficiently, @fig-intro b,c, compared to arbitrary $\Sigma_{\text{sim}}$ which typically require a Cholesky or eigenvalue decomposition.
Specifically, random samples are generated using the fact that if $u,v$ are independent vectors with iid standard normal entries, then $Dv + UWu$ has covariance matrix equal to $D^2 + UW^2U^T$, see Methods.

@@

To compare the three simulation methods with a real dataset, we chose as a references dataset 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
All data were simulated with negative binomial distributions fit by DESeq2.
We repeated the simulations, each of 12 samples, a total of 8 times to estimate variance.
For the PCA method, we used $k=2$ dimensions and for the spiked Wishart, $k=11$.
The coprcor method always uses the full data matrix, analogous to $k=11$.
Note that PCA method must use a rank $k<11$ in order to generate full-rank data, see Methods, so these parameters are not directly comparable across methods.

@@

Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 100 mean reads in the real dataset).
The simulation with independence showed the least levels of gene-gene correlations (@fig-compare-to-real d).
However, the PCA method overshot the reference dataset and the spiked Wishart and corpcor methods only slightly improved upon the independent simulation.
The SPsimSeq simulator performed similarly to the PCA method.

@@

We further demonstrated our methods on metabolomics data, where we use normal marginal distributions and with just 42 features measured, see Supplemental Figure 3.

@@

We benchmarked DESeq2 [@Love2014], a popular differential expression analysis tool, using datasets simulated with dependence and ones simulated without dependence to compare its performances on both.
DESeq2 presents an interesting case because several aspects of it assume independence of genes and so may be adversely affected by gene-gene dependence.
First, the independent filtering step [@Bourgon2010-uv] assumes independence but has been reported to be robust to typical gene-gene dependence.
Relatedly, the false discovery rate (FDR) [@BH_FDR] allows only certain forms of dependence.
Lastly, DESeq2's empirical Bayes steps could possibly be affected by gene dependence.

We used a fly whole body RNA-Seq dataset GSE81142, see Supplemental Figure 1.
To select a "control" dataset samples were selected if they met the following criteria: male flies; without treatment; and after at least 2 hours of feeding.
Unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
We then randomly selected 5% of the genes to be differentially expressed, with absolute $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down regulated chosen randomly, and simulated 5 "experimental" samples.
This was repeated 20 times for each of four dependence methods (independent, PCA, Wishart, and corpcor).

@@

We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
We chose a mouse cortex time series dataset GSE151565, see Supplemental Figure 2, which contains a total of 77 samples every 3 hours, for 36 hours.
As before, unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
We computed the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples.
We computed the means of gene expressions at each time point.
We then used these to simulate 20 time series datasets for each of the four simulation methods: independent, PCA, Wishart, and corpcor.

@@

Generating data for $\Sigma_{\text{sim}}$ of this form is highly efficient.
We use two basic facts about the multivariate distribution.
First, if $A$ is a matrix and $u \sim N(0, \Sigma)$ then $Au \sim N(0, A \Sigma A^T)$.
Second, if $u \sim N(0, \Sigma_1)$ and $v \sim N(0, \Sigma_2)$, then $u + v \sim N(0, \Sigma_1 + \Sigma_2)$.
Therefore, given matrices $D, U, W$ and $u \sim N(0, I_k)$ and $v \sim N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
$$ D v + U W u \sim N(0, D^2 + UW^2 U^T) = N(0, \Sigma) $$
as desired.

In contrast, when given an arbitrary $\Sigma$, to generate a random value in $N(0, \Sigma)$, one must compute a matrix $V$ such that $VV^T = \Sigma$.
Then, using $v \sim N(0,I)$ , we have $Vv \sim N(0, \Sigma)$ as desired.
However, computing $V$ is done using a Cholesky decomposition or eigenvector decomposition, both of which are computationally expensive when $\Sigma$ is large.

@@

Building on the multivariate normal distribution, a popular approach to describe dependence in a high-dimensional settings is called the Gaussian copula approach.
The idea of this approach is that by applying a normalizing transform and later reversing the transformation, data that does not fit a normal distribution can still have its dependence structure described using a multivariate normal distribution.
This allows the marginal (i.e., univariate) distributions of each genes to be specified separately from the dependence between genes.
This operates first by transforming each gene by fitting a distribution (such as a normal distribution, Poisson, negative binomial, or other form), and then applying the fit cumulative distribution function (CDF) to the observed values.
Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
These values are then used to compute a covariance matrix $\Sigma$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.

Here, we describe the approach using the form of covariance matrix $\Sigma = D^2 + UW^2U^T$ as above.
Once data is obtained $Z \sim N(0, \Sigma)$, then one can undo the normalizing transformation to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma$.
We describe this in detail:

1.  Fit marginal distributions to each feature in $X$ to determine CDFs $F_{i}$ for each feature.
2.  Apply normalizing transform to $X$ by setting $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (see Methods).
4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
5.  Set $z' = UWu + D v$.
6.  Output the vector $x'$ where $x'_i = F_i^{-1}(\Phi(z'))$.

@@

Below, we describe the three methods for selecting the components of the covariance matrix $\Sigma_{\text{sim}} := D^2 + U W^2 U^T$.

@@

Here we have described the well-known Gaussian copula approach and argued that a specific form of covariance matrix which is well tailored to omics data simulation.
We developed three methods using this form of covariance matrix which can be used to mimic a reference dataset for simulation.
All of these methods use a multivariate normal distribution as an intermediate step and therefore substantially restrict the kinds of dependence that can be simulated.
However, when operating in a high-dimensional space some simplification is required.

To encourage adoption of dependence in simulated omics data, we developed `dependentsimr`, an R package that generates omics-scale data with realistic correlation.
This implementation is efficient and simple, requiring just two lines of code to fit a model to a reference dataset and then simulate data from it.
We demonstrated this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions as well as a metabolomics dataset using normal marginal distributions.
Our package is even more general and supports normal, Poisson, negative binomial, and arbitrary ordered discrete distributions using the empirical CDF.
Moreover, it can support multi-modal data such as is increasingly common in multi-omics.
In addition, the techniques described here are relatively simple to implement and we hope they can see broader adoption outside of our package.

>>>>

---
title: Generating Correlated Data for Omics Simulation
author:
    - name: "Jianing Yang"
      orcid: 0000-0002-2048-9398
      affiliations:
        - name: "Chronobiology and Sleep Institute, University of Pennsylvania"
    - name: "Gregory R. Grant"
      orcid: 0000-0002-0139-7658
      affiliations:
        - name: "Department of Genetics, University of Pennsylvania"
    - name: "Thomas G. Brooks"
      orcid: 0000-0002-6980-0079
      affiliations:
        - name: "Institute for Translational Medicine and Therapeutics, University of Pennsylvania"
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  cache: true
  warning: false
format:
  docx:
    fig-dpi: 800
  html:
    toc: true
    toc-expand: true
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    margin-left: 0.75in
    margin-right: 0.75in
    margin-top: 0.75in
    margin-bottom: 0.75in
bibliography: references.bib
---
## Abstract

Simulation of realistic omics data is a key input for benchmarking studies that help users obtain optimal computational pipelines.
Omics data involves large numbers of measured features on each sample and these measures are generally correlated with each other.
However, simulation too often ignores these correlations, perhaps due to computational and statistical hurdles of doing so.
To alleviate this, we describe three approaches for generating omics-scale data with correlated measures which mimic real datasets.
These approaches are all based on a Gaussian copula approach with a covariance matrix that decomposes into a diagonal part and a low-rank part.
This decomposition allows for extremely efficient simulation, overcoming a hurdle for adoption of past methods.
We use these approaches to demonstrate the importance of including correlation in two benchmarking applications.
First, we show that variance of results from the popular DESeq2 method increases when dependence is included.
Second, we demonstrate that CYCLOPS, a method for inferring circadian time of collection from transcriptomics, improves in performance when given gene-gene dependencies in some circumstances.
We provide an R package, dependentsimr, that has efficient implementations of these methods and can generate dependent data with arbitrary marginal distributions, including discrete (binary, ordered categorical, Poisson, negative binomial), continuous (normal), or with an empirical distribution.

@@

Modern techniques, including high-throughput sequencing, produce more data than ever before.
To determine the optimal computational analysis methods for these data, benchmarks are often performed using simulated data.
This simulated data needs to closely match realistic data in order for benchmarking to meaningful.
An often neglected aspect of this is that measurements of different values are often correlated or dependent upon each other.
Two possible reasons for this neglect could be that there is a lack of guidelines on how to produce such data and also that methods to produce it are computationally expensive to run.
We describe here three related methods that are both conceptually relatively simple and also highly computationally efficient.
We demonstrated these on two applications which show how inclusion of these dependencies can affect the results of benchmarking.
Lastly, we provide a software package to act as a reference implementations of these.

@@

Using a Gaussian copula [@Nelsen1998-nq] approach (also referred to as NORTA, for "normal to anything" [@norta]), the marginal (univariate) distributions can have realistic forms.
These solutions operate by taking a real, reference dataset and producing simulations that mimic it.
However, a major challenge when simulating datasets is choice of the copula, which is a mathematical object that encodes the dependence between two or more variables.
For Guassian copulas, this is a choice of a covariance matrix.
However, the sample covariance matrix of a reference dataset is a poor choice and we present three alternative choices.

There is a large body of literature on determining covariance matrices for various purposes.
Shrinkage methods [@SchaferStrimmer2005; @OpgenRheinStrimmer2007] improve estimates and guarantee that the estimated matrix is well-conditioned (and hence invertible).
Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] produce matrices where many of the pairwise correlations are zero.
Similarly, other methods produce a sparse estimate of the inverse of the covariance matrix [@sparse_inverse_covariance], which imparts pairwise independence of some variables conditional on all the others.

Copula approaches are flexible and have a long history [@Sklar] of use in fields such as economics [@MacKenzie_copula].
More recently, copula approaches have been used for modelling or classification in multi-omics fields [@He_copula; @Ma_copula] and metagenomics [@Deek_microbiome].
Projects simulating omics with copulas include SPsimSeq [@Assefa2020] and scDesign2 [@Sun2021].
Some methods have employed the "vine copula" approach in the context of single-cell or spatial omics simulation [@fuetterer2019constructing; @scDesign3].
Vine copulas provide an alternative approach for choosing a copula which decomposes it into a set of 2-dimensional dependencies, however it comes at significant computational cost when done at the omics-scale.
SeqNet uses instead a network approach for gene dependence [@SeqNet].

Here, we describe three different strategies for determining covariance matrices, all of which enable efficient generation of simulated data due to following a specific form.
Specifically, these all decompose into a diagonal part and a low-rank part, which speeds up generation of random vectors.
First, the 'PCA' method uses principal component analysis (PCA) and picks the low-rank part such that the simulated data has the same variance in the top $k$ PCA components of the reference dataset.
Second, the 'spiked Wishart' method fits $k$ components such that simulations with the same number of samples as the reference dataset will have, on average, the same PCA component variances as the reference.
Unlike 'PCA', these variances are computed with respect to the simulated data's own PCA rather than using the PCA weights of the reference dataset.
Third, the 'corpcor' method builds off the popular R library `corpcor` [@SchaferStrimmer2005; @OpgenRheinStrimmer2007], which implements a James-Stein type shrinkage estimator for the covariance matrix as a linear interpolation of the sample covariance matrix and a diagonal matrix.
No method exactly replicates the reference dataset's properties, indicating room for future research, but all improve upon the common approach of assuming independence.
We implement this in an R package which supports normal, Poisson, DESeq2-based (negative binomial with sample-specific size factors [@Love2014]), and empirical (for ordinal data) marginal distributions.

@@

We start with a reference dataset $X$ given by a $p \times n$ data matrix of $p$ features measured in each of $n$ independent samples, which could be any real dataset of interest.
Our goal is to be able to generate simulated datasets that look the same as this reference dataset, on average and when $n$ samples are simulated.
To do this, we need to capture correlations between the $p$ features, which could represent gene expressions, protein abundances, or other measured values.
We refer to these features as genes for simplicity.

Our methods are based off the well-known Gaussian copula method, which uses a multivariate normal distribution to capture aspects of the gene-gene dependence, without necessarily assuming that the data is normally distributed.
See @fig-intro a for a high-level overview.
This multivariate normal distribution is parameterized by a $p \times p$ covariance matrix, $\Sigma$.
In omics, choosing the covariance matrix $\Sigma_{\text{sim}}$ to use during simulation presents a challenge since $n < p$.
The most obvious choice is to use the sample covariance matrix $\widehat \Sigma_{\text{ref}}$ of the reference dataset (possibly after some transformation of the data).
However, using $\widehat \Sigma_{\text{ref}}$ is a poor choice for $\Sigma_{\text{sim}}$.
Indeed, we consider a choice of $\Sigma_{\text{sim}}$ to be successful if datasets of $n$ samples simulated using $\Sigma_{\text{sim}}$ have sample covariance matrices $\widehat \Sigma_{\text{sim}}$ that are similar to $\widehat \Sigma_{\text{ref}}$.
Counter-intuitively, data simulated using $\Sigma_{\text{sim}} := \widehat \Sigma_{\text{ref}}$ have sample covariance matrices that systematically and substantially differ from $\widehat \Sigma_{\text{ref}}$, see @fig-intro b.

```{r}
#| label: fig-intro
#| fig-cap: "(a) Schematic of our Gaussian copula-based approach. The goal is to pick $\\Sigma_{\\text{sim}}$ so that $\\widehat \\Sigma_{\\text{sim}} \\approx \\widehat \\Sigma_{\\text{ref}}$ while using a special, highly efficient form. (b) Eigenvalues of the sample covariance matrix $\\widehat \\Sigma_{\\text{ref}}$ of reference data compared to $\\widehat \\Sigma_{\\text{sim}}$ from simulated data using $\\Sigma_{\\text{sim}} := \\widehat \\Sigma_{\\text{ref}}$. 100 genes and 12 samples were used for both reference and the 20 simulated datasets. Both reference and simulated data follow a multivariate normal distribution. Eigenvalues are numbered from largest to smallest. (c-d) Comparison of (c) run time and (d) memory usage for our methods (pca, wishart, coprcor), SPsimSeq, multivariate normal sampling (using the mvrnorm function from the MASS package), and vine copula (implemented by the rvinecopulib R package). 100 samples were generated with a varying number of genes. A mouse cortex dataset GSE151564 was randomly subsetted to have the required number of genes to be used as the reference data set. Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts."
#| fig-width: 8
#| fig-height: 6

@@

One reason this occurs is that $\widehat \Sigma_{\text{ref}}$ is low rank, specifically at most rank $n - 1$, which is much less than $p$ in typical omics studies.
This means that any data generated using $\Sigma_{\text{sim}} = \widehat \Sigma_{\text{ref}}$ will lie on an $n-1$-dimensional hyperplane.
Intuitively, this means that the entire probability distribution of potentially $p$ dimensional data has been squished into a $n-1$ dimensional space, generating much higher gene-gene correlations than in the reference dataset.
In contrast, we know that omics contains at least some level of random noise that is essentially independent from one gene to the next.
For example, RNA-seq involves a random sampling of fragments from the true sample of RNA molecules and other omics modes involve various background noise effects.
We therefore need to increase the amount of independence beyond what $\widehat \Sigma_{\text{ref}}$ has, a process we refer to as "injecting" independence.

We present three methods - to be called PCA, spiked Wishart, and corpcor - for making the choice of $\Sigma_{\text{sim}}$
All three take the form
$$ \Sigma_{\text{sim}} = D^2 + UW^2U^T $$
where $D$ is a diagonal matrix capturing the injected independence and $UW^2U^T$ is a low-rank correlation matrix derived from $\widehat \Sigma_{\text{ref}}$.
An additional advantage of this special form is that data may be generated very efficiently (@fig-intro b,c) compared to using arbitrary $\Sigma_{\text{sim}}$ which typically require a Cholesky or eigenvalue decomposition.
Specifically, random samples are generated using the fact that if $u,v$ are independent vectors with iid standard normal entries, then $Dv + UWu$ has covariance matrix equal to $D^2 + UW^2U^T$, see Methods.

@@

To compare the three simulation methods with a real dataset, we chose as a reference dataset 12 mouse cortex RNA-seq samples from accession GSE151923 [@Wang2022].
We then simulated data mimicking this reference using all three simulation methods (PCA, spiked Wishart, and corpcor) as well as a simulation with independent genes.
All data were simulated with negative binomial distributions fit by DESeq2.
We repeated the simulations, each of 12 samples, a total of 8 times to estimate variance.
For the PCA method, we used $k=2$ dimensions and for the spiked Wishart, $k=11$.
The coprcor method always uses the full data matrix, analogous to $k=11$.
Note that PCA method must use a rank $k<11$ in order to generate full-rank data, see Methods, so these parameters are not directly comparable across methods.

@@

Then, we computed the gene-gene correlation on pairs of high-expressed genes (at least 100 mean reads in the real dataset).
The simulation with independence showed the least levels of gene-gene correlations (@fig-compare-to-real d).
However, the PCA method overshot the reference dataset and the spiked Wishart and corpcor methods modestly improved upon the independent simulation.
The SPsimSeq simulator performed similarly to the PCA method.

@@

We further demonstrated our methods on metabolomics data, where we use normal marginal distributions and with just 42 features measured, see Supplemental Figure 1.

@@

We benchmarked DESeq2 [@Love2014], a popular differential expression analysis tool, using datasets simulated with dependence and ones simulated without dependence to compare its performance on both.
DESeq2 presents an interesting case because several aspects of it assume independence of genes and so may be adversely affected by gene-gene dependence.
First, the independent filtering step [@Bourgon2010-uv] assumes independence but has been reported to be robust to typical gene-gene dependence.
Relatedly, the false discovery rate (FDR) [@BH_FDR] allows only certain forms of dependence [@Benjamini2001-na].
Lastly, DESeq2's empirical Bayes steps could possibly be affected by gene dependence.

We used a fly whole body RNA-seq dataset GSE81142, see Supplemental Figure 2.
To select a "control" dataset samples were selected if they met the following criteria: male flies; without treatment; and after at least 2 hours of feeding.
Unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
We then randomly selected 5% of the genes to be differentially expressed, with absolute $\log_2$ fold change uniformly distributed between $0.2$ and $2.0$, either up or down regulated chosen randomly, and simulated 5 "experimental" samples.
This was repeated 20 times for each of four dependence methods (independent, PCA, Wishart, and corpcor).

@@

We next used our simulation method to benchmark CYCLOPS [@Anafi2017], which infers relative times for a set of unlabeled samples using an autoencoder to identify circular structures.
We chose a mouse cortex time series dataset GSE151565, see Supplemental Figure 3, which contains a total of 77 samples every 3 hours, for 36 hours.
As before, unnormalized read counts were used as the reference dataset and all simulations used the DESeq2-based negative binomial model.
We computed the dependence structure of the genes as well as the variances of marginal distributions using the 12 time point 0 samples.
We computed the means of gene expressions at each time point.
We then used these to simulate 20 time series datasets for each of the four simulation methods: independent, PCA, Wishart, and corpcor.

@@

Generating data for $\Sigma_{\text{sim}}$ of this form is highly efficient.
We use two basic facts about the multivariate distribution.
First, if $A$ is a matrix and $u \sim N(0, \Sigma)$ then $Au \sim N(0, A \Sigma A^T)$.
Second, if $u \sim N(0, \Sigma_1)$ and $v \sim N(0, \Sigma_2)$, then $u + v \sim N(0, \Sigma_1 + \Sigma_2)$.
Therefore, given matrices $D, U, W$ and $u \sim N(0, I_k)$ and $v \sim N(0, I_p)$ (where $I_k$ and $I_p$ are the $k \times k$ and $p \times p$ identity matrices), then
$$ D v + U W u \sim N(0, D^2 + UW^2 U^T) = N(0, \Sigma_{\text{sim}}) $$
as desired.

In contrast, when given an arbitrary $\Sigma_{\text{sim}}$, to generate a random value in $N(0, \Sigma_{\text{sim}})$, one must compute a matrix $V$ such that $VV^T = \Sigma_{\text{sim}}$.
Then, using $v \sim N(0,I)$ , we have $Vv \sim N(0, \Sigma_{\text{sim}})$ as desired.
However, computing $V$ is done using a Cholesky decomposition or eigenvector decomposition, both of which are computationally expensive when $\Sigma_{\text{sim}}$ is large.

@@

Building on the multivariate normal distribution, a popular approach to describe dependence in a high-dimensional settings is called the Gaussian copula approach.
The idea of this approach is that by applying a normalizing transform and later reversing the transformation, data that does not fit a normal distribution can still have its dependence structure described using a multivariate normal distribution.
This allows the marginal (i.e., univariate) distributions of each genes to be specified separately from the dependence between genes.
This operates first by transforming each gene by fitting a distribution (such as a normal distribution, Poisson, negative binomial, or other form), and then applying the fit cumulative distribution function (CDF) to the observed values.
Finally, those are fed to a standard normal distribution's inverse CDF to obtain values that are approximately normally distributed.
These values are then used to compute a covariance matrix $\Sigma_{\text{sim}}$ and the data is assumed to follow a multivariate normal distribution in $p$ dimensions with that covariance matrix.

Here, we describe the approach using the form of covariance matrix $\Sigma_{\text{sim}} = D^2 + UW^2U^T$ as above.
Once data is obtained $Z \sim N(0, \Sigma_{\text{sim}})$, then one can undo the normalizing transformation to obtain data with the same marginal distributions as the fit marginal distributions but with dependence determined by $\Sigma_{\text{sim}}$.
We describe this in detail:

1.  Fit marginal distributions to each feature in $X$ to determine CDFs $F_{i}$ for each feature.
2.  Apply normalizing transform to $X$ by setting $Z_{ij} = \Phi^{-1}(F_{i}(X_{ij}))$ where $\Phi$ is the CDF of the standard normal distribution.
3.  Compute $D$, $U$, $W$ matrices from $X$ by one of three methods (described later).
4.  Generate $k$ i.i.d. standard normally distributed values $u$ and $p$ i.i.d standard normally distributed values $v$.
5.  Set $z' = UWu + D v$.
6.  Output the vector $x'$ where $x'_i = F_i^{-1}(\Phi(z'))$.

@@

Below, we describe the three methods for selecting the components $D, U, W$ of the covariance matrix.

@@

Here we have described the well-known Gaussian copula approach and argued that a specific form of covariance matrix is well tailored to omics data simulation.
We developed three methods using this form of covariance matrix which can be used to mimic a reference dataset for simulation.
All of these methods use a multivariate normal distribution as an intermediate step and therefore substantially restrict the kinds of dependence that can be simulated.
However, when operating in a high-dimensional space some simplification is required.

To encourage adoption of dependence in simulated omics data, we developed `dependentsimr`, an R package that generates omics-scale data with realistic correlation.
This implementation is efficient and simple, requiring just two lines of code to fit a model to a reference dataset and then simulate data from it.
We demonstrated this package on RNA-seq data, using the DESeq2 method to fit negative binomial marginal distributions, as well as a metabolomics dataset using normal marginal distributions.
Our package is even more general and supports normal, Poisson, negative binomial, and arbitrary ordered discrete distributions using the empirical CDF.
Moreover, it can support multi-modal data such as is increasingly common in multi-omics.
In addition, the techniques described here are relatively simple to implement and we hope they can see broader adoption outside of our package.

We demonstrated that our methods are highly efficient and easily scale to the large feature counts typical of omics datasets.
While runtime of simulators is typically not critical, we posit that it introduces a sufficient inconvenience to have slowed the adoption of dependence in omics simulation and could be particularly cumbersome whenever large numbers of simulated datasets are needed.
Moreover, high memory use could be an even larger barrier than run-time, and is also dramatically improved by our methods.
Of note is that alternative methods are typically run with only a subset of genes chosen, for example, 500-2000 genes for scDesign2 [@Sun2021], 1000 genes for scDesign3 [@scDesign3], and 621 for SeqNet [@SeqNet]; presumably these limitations are due to the high resource requirements.
In contrast, our methods can operate on the entire space of genes, avoiding an arbitrary choice of cutoff, though it should be noted that correlation is most meaningful in high-expressed, high-variance genes and so subsetting genes is reasonable.
